---
title: "Assignment1_ArujanSrimohan_100750053"
author: "Arujan Srimohan"
date: "`r Sys.Date()`"
output: html_document
---

## Tracking Canada's Adoption of Electric Vehicles
With the rising concern of climate change and prices of natural resources such as 
gas and oil ballooning, the electric vehicle market is on the rise. These upcoming charts will show us how
the EV market has been changing over the years and how the adoption rate differs across the country.
This data was collected from the government of Canada called *New motor vehicle registrations, quarterly - Dataset*. 
```{r libraries, echo = FALSE, include = TRUE, message = FALSE}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(scales)
library(ggalluvial)
library(sf)
library(maps)
library(canadianmaps)

vehicle_data <- read_csv("canada_vehicle_registration.csv")

#First we will drop all the unwanted vehicle data to work with a more concise
#data set
vehicle_data <- filter(vehicle_data,
    `Fuel type` %in% c("Gasoline", "Hybrid electric",
                        "Plug-in hybrid electric", "Battery electric",
                        "All zero-emission vehicles"))

#We also need to remove a couple provinces that don't share data
vehicle_data <- filter(vehicle_data,
    !GEO %in% c("Alberta", "Newfoundland and Labrador", "Nova Scotia"))

#Then we will combine repetitive categories to help our dataset
#retain focus on the topic
vehicle_data <- mutate(vehicle_data, `Fuel Category` = case_when(
    `Fuel type` == "Gasoline" ~ "Gasoline Vehicles",
    `Fuel type` %in% c("Hybrid electric",
                       "Plug-in hybrid electric") ~ "Hybrid Vehicles",
    `Fuel type` %in% c("Battery electric",
                       "All zero-emission vehicles") ~ "Electric Vehicles"),
    #Convert to numeric to use as continuous variable
    Year = as.numeric(substr(REF_DATE, 1, 4)))

# The dataset will have rows with repetitive new groupings so lets clean them up
vehicle_data <- group_by(vehicle_data,
                         Year, GEO, `Vehicle type`, `Fuel Category`) |>
    summarize(registrations = sum(VALUE))

#vehicle_data is now our base cleaned data
```

First, let's examine the share each fuel type has on the market at the beginning of 2025:
```{r, echo = FALSE, fig.width=12, fig.height=6, out.width = "100%", include = TRUE, message = FALSE}
#Let's use all the "toal, vehicle type" counts for canada in 2024
canada_totals <- filter(vehicle_data, `Vehicle type` == "Total, vehicle type",
    Year == 2024,
    GEO == "Canada")

#Then let's calculate the share each count has against the sum of all counts
canada_totals <- mutate(canada_totals,
                        share = registrations / sum(registrations))

#Use a pie chart to show proportions using share column we calculated
#and fill using the registrations per fuel category
ggplot(canada_totals, aes(x = "", y = registrations, fill = `Fuel Category`,
                          label = percent(share))) +
    geom_bar(stat = "identity", width = 1) +
    geom_text(position = position_stack(vjust = 0.5)) +
    coord_polar(theta = "y") +
    scale_fill_brewer(
        palette = "RdYlBu",
        name = "Fuel Category"
    ) +
    theme_void() +
    labs(title = "Vehicle Registrations by Fuel Category 2024",
        color = "Fuel Category"
    ) +
    theme(plot.title = element_text(face = "bold", size= 20, hjust = 0.5),
          legend.position = "left",
          legend.title = element_text(size = 15),
          legend.text = element_text(face = "italic", size = 12))
```
This pie chart provides a clear view of how the market is divided among the categories at a national 
level. Currently, eco-friendly vehicle options are at an adoption rate of about **35%** across the country.

What does this vehicle share look like across the country?
```{r, echo = FALSE, fig.width=12, fig.height=6, out.width = "100%", include = TRUE, message = FALSE}
#Take our base cleaned data and remove national data rows and rows not in 2024
#and keep the totals for vehicle type data
fuel_type_totals <- filter(vehicle_data, `Vehicle type` == "Total, vehicle type",
    Year == 2024,
    GEO != "Canada")

#Now plot a grouped bar chart, with each group being a province, and 3 bars per group
#where the bars are the fuel category
ggplot(fuel_type_totals, aes(x = GEO, y = registrations, fill = `Fuel Category`)) +
    geom_bar(stat = "identity", position = position_dodge()) +
    scale_y_continuous(breaks = breaks_width(25000),
                       labels = comma,
                       expand = c(0, 0)) +
    labs(title = "Provincial Vehicle Registrations by Fuel Category 2024",
        x = "Province",
        y = "Number of Vehicles Registered",
        color = "Fuel Category"
    ) +
    scale_fill_manual(
        values = c("Gasoline Vehicles" = "firebrick3",
                "Hybrid Vehicles" = "seagreen3",
                "Electric Vehicles" = "dodgerblue1")
    ) +
    theme(plot.title = element_text(face = "bold", size= 20, hjust = 0.5),
          panel.background = element_rect(fill = "white", color = "black"),
          panel.grid.major = element_line(color = "gray", linewidth = 0.5),
          panel.grid.minor = element_blank(),
          legend.title = element_text(size = 15),
          legend.text = element_text(face = "italic", size = 12))

```
Here we compare each province's vehicle share side by side, showing us the magnitude in differences
in each fuel type's adoption. This visualization helps highlight provinces where EV registrations only
contribute a small amount to the market, and subsequently suggests that this is just a reflection of the
low registrations in the province as a whole.

Next, let's try to understand how Canada has grown to this point:
```{r, echo = FALSE, fig.width=12, fig.height=6, out.width = "100%", include = TRUE, message = FALSE}
#Filter out any row that does not have totals for vehicle type, we want
#values where the fuel category is EVs and we are look at provincial totals
#over the year not national
ev_totals <- filter(vehicle_data, `Vehicle type` == "Total, vehicle type",
    `Fuel Category` == "Electric Vehicles",
    GEO != "Canada")

#Plot the totals for the provinces over the years to see how the 
#adoption of EVs has grown
ggplot(ev_totals, aes(x = Year, y = registrations, color = GEO)) +
    geom_line(linewidth = 1) +
    geom_point(size = 2) +
    scale_x_continuous(breaks = c(2017, 2018, 2019, 2020, 2021,
                                  2022, 2023, 2024)) +
    scale_y_continuous(breaks = breaks_width(25000),
                       labels = comma) +
    labs(title = "EV Adoption over the Years",
        x = "Year",
        y = "Number of EVs Registered",
        color = "Province"
    ) +
    theme(plot.title = element_text(face = "bold", size= 20, hjust = 0.5),
          axis.line = element_line(colour = "black"),
          legend.position = "left",
          legend.title = element_text(size = 15),
          legend.text = element_text(face = "italic", size = 12))

```
If we take a look at our graph, one key consistency is the growth in EV registrations after the year 2019.
On **May 1st, 2019** the *iZEV* program was introduced by the government of Canada 
to give citizens a **$5000** rebate for any zero emission vehicle purchase.
It's clear the impact incentives can have, and this has helped mitigate the higher barrier of entry EVs pose when
trying to purchase a vehicle.  

Subsequently, we see the provinces Saskatchewan, PEI, and others unable to scratch  **10,000** 
EV registrations per year. We see highly populated provinces like Ontario and British Columbia having their growth wane.
Maybe their vehicle market share has hit a plateau. Let's take a look:
```{r, echo = FALSE, fig.width=22, fig.height=16, out.width = "100%", include = TRUE, message = FALSE}
#We want to map each fuel category to each vehicle type to each province
#Start by filtering for specific vehicle types across the provinces
vehicle_flow_data <- filter(vehicle_data, `Vehicle type` != "Total, vehicle type",
                      Year == 2024,
                      GEO != "Canada")
#Then group the results by vehicle type, fuel category and GEO, so we can summarize
#and count the total registrations for each group
vehicle_flow_data <- group_by(vehicle_flow_data,
                              `Vehicle type`, `Fuel Category`, GEO) |>
    summarize(registrations = sum(registrations))

#Create the Sankey graph that will map our areas of focus, 
#fuel categories -> vehicle type
ggplot(vehicle_flow_data, aes(y = registrations, axis2 = `Vehicle type`,
                              axis1 = `Fuel Category`)) +
  #Alluvium flows are the band, alpha is the transparency
  geom_alluvium(aes(fill = `Fuel Category`), width = 1/10, alpha = 0.5) +
  #Stratum is our axis labelling method for the sankey(label blocks)
  geom_stratum(width = 1/10, fill = "grey", color = "black") +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)),
            angle = -15, hjust = 0.4, size = 7) +
  #Make the plot narrower  
  scale_x_discrete(expand = c(0.1, 0.1)) +
  scale_fill_manual(
    values = c("Gasoline Vehicles" = "firebrick3",
                "Hybrid Vehicles" = "seagreen3",
                "Electric Vehicles" = "dodgerblue1")
  ) +
  #Need more space because of long label names
  coord_cartesian(clip = "off") +
  labs(
    title = "Flow of Vehicle Registrations to Type 2024",
  ) +
  theme_void(base_size = 20) +
  theme(plot.title = element_text(face = "bold", size= 30, hjust = 0.5),
        legend.title = element_text(size = 25),
        legend.text = element_text(face = "italic", size = 20))

#Lets make the sankey to complete our mapping 
#vehicle type -> province
ggplot(vehicle_flow_data, aes(y = registrations, axis2 = `Vehicle type`,
                              axis1 = `GEO`)) +
  geom_alluvium(aes(fill = `Fuel Category`), width = 1/10, alpha = 0.5) +
  geom_stratum(width = 1/10, fill = "grey", color = "black") +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)),
            angle = -15, hjust = 0.4, size = 7) +
  scale_fill_manual(
    values = c("Gasoline Vehicles" = "firebrick3",
                "Hybrid Vehicles" = "seagreen3",
                "Electric Vehicles" = "dodgerblue1")
  ) +
  #Make vertical flowing sankey
  coord_flip(clip = "off") +
  labs(
    title = "Flow of Vehicle Type to Provinces Registered in 2024",
  ) +
  theme_void(base_size = 20) +
  theme(plot.title = element_text(face = "bold", size= 30, hjust = 0.5),
        legend.title = element_text(size = 25),
        legend.text = element_text(face = "italic", size = 20))
```
The first sankey plot shows us which fuel categories are most registered in which vehicle types. The second plot
shows us the volume of the vehicle types being registered in each province. Understanding the vehicle 
segments driving EV adoption is critical to see where improvements are needed. These diagrams show us
the success of multi-purpose vehicles, particularly in Quebec, where EV adoption now surpasses
gasoline options. Conversely, we find that bigger vehicles with more emissions are harder to electrify well because of 
factors like **mileage, horsepower, and torque**. This highlights the impact that technological readiness and suitability has
on the adoption of EVs. 

Finally, let's look at the growth of EVs over time from a geographical context.
```{r, echo = FALSE, fig.width=12, fig.height=8, out.width = "100%", include = TRUE, message = FALSE}
#Take our base cleaned data and remove national data rows and rows not in 2024
#and keep the Totals for vehicle type data per province
province_vehicle_totals <- filter(vehicle_data,
                                  `Vehicle type` == "Total, vehicle type",
                                  Year %in% c(2017,2024),
                                  GEO != "Canada")

#Then group the results by GEO, so we can summarize
#and count the total registrations for each province, there is a catch though
province_vehicle_totals <- group_by(province_vehicle_totals, Year, GEO) |>
    summarize(
        total_registrations = sum(registrations),
        ev_registrations = sum(ifelse(`Fuel Category` == "Electric Vehicles",
                                      registrations, 0)))

#Get a percentage that we will map
province_vehicle_totals <- province_vehicle_totals %>%
  mutate(ev_percentage = (ev_registrations / total_registrations) * 100)

#We need to make rows for the territories since the dataset combines them all
#with BC, let's clone the BC row and mutate it into each territory
#this approach will let BC and all the territories have the same shared value

#Extract bc data
bc_territory_rows <- province_vehicle_totals %>%
  filter(GEO == "British Columbia and the Territories")

#Create extra rows for each province/territory and mutate to change
#the name
bc_row <- mutate(bc_territory_rows, GEO = "British Columbia")
yukon_row <- mutate(bc_territory_rows, GEO = "Yukon")
nwt_row <- mutate(bc_territory_rows, GEO = "Northwest Territories")
nunavut_row <- mutate(bc_territory_rows, GEO = "Nunavut")

# Combine the extra rows
extra_rows <- bind_rows(bc_row, yukon_row, nwt_row, nunavut_row)

# glimpse(extra_rows)
# Remove the combined row from the original dataset and add the new rows
province_vehicle_totals_final <- filter(province_vehicle_totals,
                                        GEO != "British Columbia and the Territories")

province_vehicle_totals_final <- bind_rows(province_vehicle_totals_final,
                                           extra_rows)


#Now lets combine the geometry data from canadianmaps to our registrations
#The prename column is equivalent to GEO if we look at candianmaps docs
canada_data <- left_join(PROV, province_vehicle_totals_final,
                         join_by("PRENAME" == "GEO"))
#Remove all null data points because left join will retain
#data for all provinces in canadianmaps but our dataset
#wasn't allowed to collect Alberta, Newfoundland and Labrador, Nova Scotia
canada_data <- canada_data |> filter(!is.na(Year))

#Create the sf map to show EV Registrations across the country
ggplot(canada_data, aes(fill = ev_percentage/100)) +
  geom_sf(color = "gray90", size = 0.1) +
  #Gradient for a smooth transition over our map
  scale_fill_gradient(low = "white",
    high = "midnightblue",
    labels = percent,
    name = "EV Registrations"
  ) +
  labs(title = "Electric Vehicle Registrations by Province (2024)") +
  facet_wrap(~Year)+
  theme_map() +
  #Change legend to a gradient bar
  guides(fill = guide_colorbar(theme =
                                theme(legend.key.width = unit(1, "null"),
                                      legend.frame =
                                      element_rect(colour = "black")))) +
  theme_void(base_size = 10) +
  theme(legend.position = "bottom",
        legend.title = element_text(size = 12, vjust = 0.8),
        legend.text = element_text(size = 15),
        plot.title = element_text(face = "bold", size= 20, vjust = 3, hjust = 0.5))
```
This map visualizes the percentage of EV registrations across the provinces and territories, 
offering a spatial perspective on the regional disparities we have been exploring. Provinces and territories 
where the EVs are starting to eat up market share are darker on the color gradient. Conversely, the color gradient
highlights the area with lower registrations as lighter colors, giving a clear picture of where additional 
EV infrastructure and incentives might be needed.

Overall we can conclude that while EV registrations are growing nationwide, the pace and nature of their
growth vary by region and vehicle type. Factors like charging infrastructure, government incentives,
fuel economy, and time in the market all contribute to the challenges of adopting EVs in different regions.
